---
title: "Getting started with cfbseedR"
description: "Compute standings, playoff seeds, and run a small season simulation."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cfbseedR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

cfbseedR simulates and evaluates college football seasons: standings with a
documented tiebreaker cascade, conference champions, College Football
Playoff (CFP) seeding, and week-by-week season simulation with a pluggable
results generator.

Everything in this article runs offline using the toy season bundled with
the package (9 teams: two 4-team conferences, "Alpha" and "Beta", plus one
independent).

```{r setup}
library(cfbseedR)

games <- read.csv(system.file("extdata", "toy_games.csv", package = "cfbseedR"))
teams <- read.csv(system.file("extdata", "toy_teams.csv", package = "cfbseedR"))

head(games)
teams
```

## Standings from a games frame

`cfb_standings()` needs a games frame (`sim`/`season`, `game_type`, `week`,
`home_team`, `away_team`, `result`) and a teams frame (`team`,
`conference`). It returns one row per team with overall and conference
records, conference ranks (via the tiebreaker cascade), and conference
champions.

```{r standings}
standings <- cfb_standings(games, teams, verbosity = "NONE")
standings[, c("team", "conference", "wins", "losses", "conf_pct",
              "conf_rank", "conf_champ")]
```

Two college-football-specific semantics to be aware of:

* **Conference championship games** (`game_type == "CONF_CHAMP"`) count
  toward the overall record and decide the conference champion, but not
  toward the conference record/rank.
* **`sov` / `sos` are conference-scoped**: strength of victory and strength
  of schedule are computed over regular-season conference games only
  (`sov` over conference victories, `sos` over conference opponents);
  independents get `0.0`.

If you have real data, `cfb_games_from_schedule()` maps a
`cfbfastR::load_cfb_schedules()` frame into this schema:

```{r cfbfastR, eval = FALSE}
# Not run: requires network access via cfbfastR
sched <- cfbfastR::load_cfb_schedules(2024)
games_2024 <- cfb_games_from_schedule(sched)
```

## Playoff seeds with a rankings frame

`cfb_playoff_seeds()` implements CFP **straight seeding** (2025 rule): the
field is the best-ranked teams with the 5 highest-ranked conference
champions guaranteed inclusion, seeded strictly in ranking order. Pass a
committee-style rankings frame (`team`, `rank`); without one, a documented
fallback ordering (win pct, SOV, SOS, point differential) is used.

```{r seeds}
rankings <- data.frame(team = c("B1", "I1", "A1", "A3"), rank = 1:4)
seeded <- cfb_playoff_seeds(standings, rankings = rankings, playoff_seeds = 4)
seeded[!is.na(seeded$seed), c("team", "conference", "conf_champ", "seed")]
```

## A small simulation

`cfb_simulations()` fills in the missing (`NA`) results week by week using
a `compute_results` function - by default `cfbseedR_compute_results()`, an
ELO-based generator adapted from nflseedR - then computes standings,
champions, seeds, and (with `sim_include = "POST"`) the playoff bracket.

```{r simulation}
games$result[games$week >= 3] <- NA
set.seed(42)
sim <- cfb_simulations(games, teams, simulations = 50, playoff_seeds = 4)
sim$overall
```

The returned `cfbseedR_simulation` list also carries per-simulation
`standings`, all simulated `games`, `team_wins` win-total probabilities,
per-matchup `game_summary`, and the `sim_params` used.

For deeper topics - custom `compute_results` functions, the
`tiebreaker_depth` ladder, and the CFP-12 seeding rules - see the
[Simulating seasons](https://cfbseedR.sportsdataverse.org/articles/simulating-seasons.html)
article.

## Acknowledgments

cfbseedR is an adaptation of
[nflseedR](https://nflseedr.com) (MIT) by **Lee Sharpe** and **Sebastian
Carl**. The entire architecture - the standings engine, the tiebreaking
cascade design, the week-loop simulator with a pluggable `compute_results`
contract, and the ELO-based default generator - is their design, re-derived
here with college football semantics. Thank you to Lee, Sebastian, and the
broader [nflverse](https://nflverse.nflverse.com) team whose open-source
work this package builds on.
