---
title: "Simulating seasons"
description: "Custom compute_results functions, the tiebreaker ladder, and CFP-12 seeding."
author: "Saiem Gilani <br><a href='https://x.com/saiemgilani' target='blank'><img src='https://img.shields.io/twitter/follow/saiemgilani?color=blue&amp;label=%40saiemgilani&amp;logo=x&amp;style=for-the-badge' alt='@saiemgilani'/></a> <a href='https://github.com/saiemgilani' target='blank'><img src='https://img.shields.io/github/followers/saiemgilani?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@saiemgilani'/></a>"
opengraph:
  image:
    src: "https://github.com/sportsdataverse/cfbfastR-data/blob/main/themes/social_card_cfbseedR.png?raw=true"
  twitter:
    card: summary_large_image
    creator: "@saiemgilani"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article goes deeper than the
[Getting started](https://cfbseedR.sportsdataverse.org/articles/getting-started.html)
walkthrough: writing a custom `compute_results` function, controlling the
tiebreaker cascade with `tiebreaker_depth`, and how the 12-team CFP
straight-seeding rules are applied. Everything runs offline on the bundled
toy season.

```{r setup}
library(cfbseedR)

games <- cfb_games_example
teams <- cfb_teams_example
games$result[games$week >= 3] <- NA
```

## Custom `compute_results` functions

`cfb_simulations()` calls `compute_results(teams, games, week_num, ...)`
once per remaining week. The contract (identical to nflseedR's):

* return `list(teams = teams, games = games)`,
* fill `result` for exactly the games with
  `week == week_num & is.na(result)`,
* never modify any other result, remove rows/columns from `games`, or
  produce a tie (`result == 0`) outside the regular season.

Here is a deliberately simple generator: every missing result is a home win
by 3 (postseason-safe since 3 is never a tie):

```{r custom-fn}
home_field_rules <- function(teams, games, week_num, ...) {
  fill <- games$week == week_num & is.na(games$result)
  games$result[fill] <- 3
  list(teams = teams, games = games)
}
```

`simulations_verify_fct()` checks a custom function against the full
contract before you burn simulation time on it:

```{r verify}
simulations_verify_fct(home_field_rules)
```

Then plug it in:

```{r custom-sim}
set.seed(1)
sim <- cfb_simulations(
  games, teams,
  compute_results = home_field_rules,
  simulations = 10, playoff_seeds = 4
)
sim$overall[, c("team", "wins", "conf_champ", "playoff")]
```

Anything extra your generator needs (a ratings vector, market lines, an
externally trained model) can be passed through the `...` of
`cfb_simulations()`. The default generator accepts initial ELO ratings that
way:

```{r elo-passthrough}
elo <- setNames(seq(1700, 1300, length.out = nrow(teams)), teams$team)
set.seed(2)
sim_elo <- cfb_simulations(
  games, teams,
  simulations = 10, playoff_seeds = 4,
  elo = elo
)
sim_elo$overall[, c("team", "wins", "playoff")]
```

## The `tiebreaker_depth` ladder

Conference ranks are seeded by conference win percentage; ties are broken
by a cascade whose depth you control:

| `tiebreaker_depth` | Cascade applied before the coin flip |
|---|---|
| `"RANDOM"` | Nothing - coin flip immediately. |
| `"PRE-SOV"` | Head-to-head, then common conference opponents. |
| `"SOS"` (default) | `"PRE-SOV"` + strength of victory, then strength of schedule. |
| `"POINTS"` | `"SOS"` + conference point differential. |

`verbosity = "MAX"` logs every step the cascade takes, which is the easiest
way to understand (and trust) a rank:

```{r tiebreakers}
played <- cfb_games_example
standings <- cfb_standings(played, teams,
                           tiebreaker_depth = "POINTS", verbosity = "MAX")
standings[, c("team", "conference", "conf_pct", "sov", "sos", "conf_rank")]
```

### The conference-scoped `sov`/`sos` ruling

All cascade quantities - head-to-head, common opponents, `sov`, `sos`, and
point differential - are computed **over regular-season conference games
only**, so conference ranks depend only on conference play. Concretely:

* `sov` = beaten conference opponents' conference wins divided by their
  conference games (strength of victory over **conference victories**),
* `sos` = all conference opponents' conference wins divided by their
  conference games (strength of schedule over **conference opponents**),
* independents have no conference games, so both are `0.0`.

This is a documented simplification: real CFB tiebreakers are
conference-specific, and a full-schedule SOS would let non-conference
results leak into conference seeding.

### Official per-conference tiebreakers

The generic cascade above is the fallback used for conferences without a
registered procedure. The **SEC, Big Ten, Big 12, ACC, and MAC** instead use
their **official 2024+ tiebreaker procedures** (ported from the same
registry as sdv-py's `cfb_standings.py`, so both engines agree on a shared
cross-language fixture). These add rungs the generic cascade doesn't have -
record vs. common opponents by order of finish, a pooled conference
opponents' win percentage, the SEC's capped relative scoring margin (points
scored capped at 42 / allowed at 48 per game), the Big 12's `total_wins`
with its FCS-or-lower win cap, and an external analytics rating - and they
always run to completion (`tiebreaker_depth` only gates the generic
fallback). Two optional inputs feed these rungs when available:
`home_points`/`away_points` on `games` (the SEC cap) and a `division`
column on `teams` (the Big 12 FCS cap); `tiebreaker_data =
list(analytics_ratings = ratings)` supplies the external rating rung. Any
rung whose input is missing is skipped, and the skip is recorded in
`attr(standings, "tiebreak_notes")`:

```{r registry}
sec_teams <- data.frame(
  team = c("A", "B", "C"), conference = "SEC", division = "FBS"
)
sec_games <- data.frame(
  sim = 2024, week = 1:3, game_type = "REG",
  home_team = c("A", "B", "C"), away_team = c("B", "C", "A"),
  result = c(50, 3, 3), neutral = 0,
  home_points = c(50, 20, 20), away_points = c(0, 17, 17)
)
sec_standings <- cfb_standings(sec_games, sec_teams,
                               tiebreaker_depth = "POINTS", verbosity = "NONE")
sec_standings[, c("team", "conf_rank")]
attr(sec_standings, "tiebreak_notes")
```

## CFP-12 seeding and the `autobid` policy

`cfb_playoff_seeds()` always uses **straight seeding** - the field is
ordered strictly by committee rank, and champions are never bumped up to
the top-4 seeds (that was the pre-2025 rule). What changes between seasons
is *who is guaranteed a place*, which is the `autobid` argument:

* `autobid = "2026"` (the default, today's rule): the **ACC, Big 12, Big
  Ten and SEC champions are in regardless of their ranking**, the
  **highest-ranked Group-of-6 team** is in whether or not it won its
  conference, and **Notre Dame** is in when ranked inside the field.
* `autobid = "2025"`: the 5 highest-ranked **conference champions** are
  guaranteed (fewer if fewer champions exist) - the 2024-2025 rule.

Under either policy the remaining places go to the best-ranked at-large
teams, and the whole field is then seeded in rank order.

A synthetic 16-team example with 6 conference champions shows the
guarantee and the straight ordering; champion `F1` (ranked 14th) displaces
the lowest at-large team but keeps its rank-order seed:

```{r cfp12}
st <- data.frame(
  sim = 1,
  team = sprintf("%s1", LETTERS[1:16]),
  conference = c(LETTERS[1:6], LETTERS[1:10]),
  conf_champ = c(rep(TRUE, 6), rep(FALSE, 10)),
  win_pct = seq(1, 0.4, length.out = 16),
  sov = 0.5, sos = 0.5, pd = 100
)
rankings <- data.frame(
  team = st$team,
  rank = c(1, 2, 3, 5, 8, 14, 4, 6, 7, 9, 10, 11, 12, 13, 15, 16)
)
seeded <- cfb_playoff_seeds(st, rankings = rankings, playoff_seeds = 12)
seeded[!is.na(seeded$seed), c("team", "conf_champ", "seed")] |>
  (\(x) x[order(x$seed), ])()
```

In `cfb_simulations()`, the bracket generated from these seeds is the fixed
CFP-12 bracket (first round hosted by the higher seed, later rounds
neutral-site, no reseeding), and committee `rankings` are held static
across simulations.

## Running simulations in parallel

Simulations are split into `chunks` (default 8) and dispatched with furrr,
so a parallel plan spreads them across cores:

```{r parallel, eval = FALSE}
future::plan("multisession")
sim <- cfb_simulations(games, teams, simulations = 10000, chunks = 8)
future::plan("sequential")
```

Progress reporting comes from progressr - switch it on once and every call
reports:

```{r progress, eval = FALSE}
progressr::handlers(global = TRUE)
```

A given seed reproduces exactly for a fixed `chunks`, and a parallel plan
gives identical results to a sequential one. The RNG stream does depend on
the chunking, though, so the same seed with a different `chunks` gives
different (equally valid) simulations.

## Summarising a simulation, and checking it

`summary()` renders the `overall` table as a gt table grouped by
conference, with probabilities formatted by `fmt_pct_special()` so a
long-shot reads `<1%` rather than a flat `0%` and a near-lock `>99.9%`
rather than a false `100%`. Here it is on a real 1,000-season run:

```{r summary}
set.seed(2026)
sim <- cfb_simulations(games, teams, simulations = 1000, playoff_seeds = 4,
                       chunks = 4, verbosity = "NONE")
summary(sim)
```

Before trusting any of those numbers, check that the simulation is
internally consistent. Three identities have to hold by construction, and
they are cheap to assert:

```{r sim-checks}
checks <- data.frame(
  Check = c("playoff probabilities sum to the seed count",
            "national-title probabilities sum to one champion",
            "conference titles sum to one per conference"),
  Expected = c(sim$sim_params$playoff_seeds, 1, 2),
  Observed = round(c(sum(sim$overall$playoff),
                     sum(sim$overall$won_natty),
                     sum(sim$overall$conf_champ)), 3)
)
checks$OK <- checks$Expected == checks$Observed
checks
```

All three hold exactly on this run. When one drifts it is nearly always a
data problem rather than a simulation problem: a team appearing in `games`
but missing from `teams`, a conference with no `CONF_CHAMP` fixture, or a
custom `compute_results` that left some scheduled results as `NA`.

The per-team win distribution is the other output worth eyeballing, since
it exposes the shape of a season rather than just its mean:

```{r team-wins}
tw <- sim$team_wins[sim$team_wins$team %in% c("A1", "B1"), ]
tw <- tw[tw$wins %in% c(1, 2, 3, 4), c("team", "wins", "over_prob")]
tw
```

## Real schedules

With cfbfastR installed, a real season plugs straight in:

```{r real, eval = FALSE}
# Not run: requires network access via cfbfastR
sched <- cfbfastR::load_cfb_schedules(2026)
real_games <- cfb_games_from_schedule(sched)
real_games <- real_games[real_games$game_type != "POST", ]
real_teams <- unique(data.frame(
  team = c(sched$home_team, sched$away_team),
  conference = c(sched$home_conference, sched$away_conference)
))

future::plan("multisession")
sim <- cfb_simulations(real_games, real_teams, simulations = 10000, chunks = 8)
future::plan("sequential")

summary(sim)
```

The same three identities apply to a real season and are worth asserting
there too — with 130-odd FBS teams and a schedule pulled from a live API,
a missing conference label or an unplayed fixture is far likelier than in
the toy data, and the sums will say so immediately.

Because the tiebreaker registry is season-scoped, feeding a `season` column
of 2024 or 2025 ranks those seasons under the rules that were in force at
the time, while `autobid = "2025"` / `"2026"` selects the CFP era
independently. The [tiebreakers cookbook](https://cfbseedR.sportsdataverse.org/articles/tiebreakers-and-cfp.html)
works through what changes when you flip each of them.


# **Our Authors**

-   [Saiem Gilani](https://x.com/saiemgilani)
<a href="https://x.com/saiemgilani" target="blank"><img src="https://img.shields.io/twitter/follow/saiemgilani?color=blue&label=%40saiemgilani&logo=x&style=for-the-badge" alt="@saiemgilani" /></a>
<a href="https://github.com/saiemgilani" target="blank"><img src="https://img.shields.io/github/followers/saiemgilani?color=eee&logo=Github&style=for-the-badge" alt="@saiemgilani" /></a>

## **Adapted from nflseedR (they're awesome)**

cfbseedR is a college-football adaptation of
[nflseedR](https://nflseedr.com) (MIT). The standings-to-seeds architecture,
the week-loop simulator with a pluggable `compute_results`, and the ELO
default generator are all theirs.

-   [Sebastian Carl](https://x.com/mrcaseb)
<a href="https://x.com/mrcaseb" target="blank"><img src="https://img.shields.io/twitter/follow/mrcaseb?color=blue&label=%40mrcaseb&logo=x&style=for-the-badge" alt="@mrcaseb" /></a>
<a href="https://github.com/mrcaseb" target="blank"><img src="https://img.shields.io/github/followers/mrcaseb?color=eee&logo=Github&style=for-the-badge" alt="@mrcaseb" /></a>

-   [Lee Sharpe](https://x.com/LeeSharpeNFL)
<a href="https://x.com/LeeSharpeNFL" target="blank"><img src="https://img.shields.io/twitter/follow/LeeSharpeNFL?color=blue&label=%40LeeSharpeNFL&logo=x&style=for-the-badge" alt="@LeeSharpeNFL" /></a>

## **Citation**

To cite cfbseedR in publications, use:

```
Saiem Gilani (2026). cfbseedR: The SportsDataverse's R Package to Simulate
and Evaluate College Football Seasons.
https://cfbseedR.sportsdataverse.org/
```

```r
citation("cfbseedR")
```

If you use the season-simulation methodology, please also cite
[nflseedR](https://nflseedr.com).

## **Related SportsDataverse packages**

-   [**cfbfastR**](https://cfbfastR.sportsdataverse.org/) — college football
    play-by-play, schedules and rosters; feeds `cfb_games_from_schedule()`
-   [**cfb4th**](https://cfb4th.sportsdataverse.org/) — fourth-down decision
    modeling
-   [**cfbplotR**](https://cfbplotR.sportsdataverse.org/) — team logos and
    plotting helpers for ggplot2
-   [**recruitR**](https://github.com/sportsdataverse/recruitR) — recruiting data
-   [**sportsdataverse-R**](https://r.sportsdataverse.org/) — the meta-package
-   [**sportsdataverse-py**](https://py.sportsdataverse.org/) — the Python
    mirror, whose `cfb_standings` shares this package's tiebreaker rulings

A printable [cheat sheet
(PDF)](https://sportsdataverse.org/cheatsheets/cfbplotR-cfb4th-cfbseedR.pdf)
covers cfbplotR, cfb4th and cfbseedR together.
