---
title: "Cookbook: tiebreakers & the CFP, past / present / future"
description: "Worked scenarios for the official per-conference tiebreakers and the CFP automatic-qualifier policies: re-examining a settled season under new rules, building today's 12-team field, and stress-testing a 16-team one."
author: "Saiem Gilani <br><a href='https://x.com/saiemgilani' target='blank'><img src='https://img.shields.io/twitter/follow/saiemgilani?color=blue&amp;label=%40saiemgilani&amp;logo=x&amp;style=for-the-badge' alt='@saiemgilani'/></a> <a href='https://github.com/saiemgilani' target='blank'><img src='https://img.shields.io/github/followers/saiemgilani?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@saiemgilani'/></a>"
opengraph:
  image:
    src: "https://github.com/sportsdataverse/cfbfastR-data/blob/main/themes/social_card_cfbseedR.png?raw=true"
  twitter:
    card: summary_large_image
    creator: "@saiemgilani"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

cfbseedR implements each FBS conference's **official championship-game
tiebreaker procedure** as season-scoped epochs, and the CFP's
**season-keyed automatic-qualifier policies**. That combination lets you
ask the question the rulebook keeps forcing on college football: *would
last year's outcome still hold under this year's rules?*

This cookbook works through three time horizons — a settled season
re-examined, the present field, and a plausible future one — and reports
what actually comes out.

```{r setup, message = FALSE}
library(cfbseedR)
library(gt)
```

A small helper so every table below renders the same way:

```{r helper}
show <- function(x, title, subtitle = NULL) {
  gt::gt(x) |>
    gt::tab_header(title = title, subtitle = subtitle) |>
    gt::tab_options(
      table.font.size = gt::px(13),
      column_labels.font.weight = "bold",
      data_row.padding = gt::px(4)
    )
}
```

> **On the data below.** The team names and records are *constructed* to
> isolate one rule difference at a time — they are not transcribed from a
> real box score. The **rules** are real, taken from the conferences' and
> the CFP's published policies. Every table is the package's actual output,
> and the last section shows how to point the same code at a real season
> through cfbfastR.

## The rules, at a glance (2026 season)

```{r rules}
rules <- data.frame(
  Conference = c("SEC", "Big Ten", "Big 12", "ACC (2026+)", "American",
                 "Conference USA", "MAC", "Mountain West", "Sun Belt",
                 "Pac-12 (re-formed)"),
  Format = c(rep("single table", 8), "two divisions", "single table"),
  `After win pct` = c(
    "h2h -> common -> descent -> opp win% -> capped margin (42/48) -> draw",
    "h2h -> common -> descent -> opp win% -> analytics -> draw",
    "h2h -> common -> descent -> opp win% -> capped wins -> analytics -> draw",
    "candidate pool (wins OR losses) -> h2h -> SportSource SSR -> draw",
    "h2h -> CFP-ranked final week -> metric composite -> common -> draw",
    "h2h -> CFP-ranked final week -> composite -> APR -> draw",
    "h2h -> common opponents -> SportSource rating -> draw",
    "metric composite -> h2h -> draw",
    "h2h -> div win% -> descent -> common -> CFP clause -> composite -> draw",
    "no published procedure: generic fallback"
  ),
  check.names = FALSE
)
show(rules, "Registered tiebreaker procedures",
     "Unregistered conferences use the documented generic cascade")
```

## PAST — re-examining a settled tie

The ACC replaced its entire tiebreaker in July 2026 for the move to a
nine-game schedule. The new policy is short — head-to-head, then the
SportSource Team Success Ranking, then a draw — but it adds a **candidate
pool** rule: a team that played a different number of conference games
joins the tie if it matches the leaders' win count *or* loss count.

That pool rule is the part that rewrites history. Here is one league's
season ranked twice, from **identical games** — once as a 2025 season,
once as 2026:

```{r past-setup}
acc_teams <- data.frame(
  team = c("Ashwood", "Brightwater", "Cedarcrest", "Dunmore", "Eastvale",
           "Fairmont", "Glenrock", "Harborview", "Ironside"),
  conference = "ACC"
)
mk <- function(season, w, h, a, r) {
  data.frame(season = season, week = w, game_type = "REG",
             home_team = h, away_team = a, result = r)
}
acc_games <- function(season) rbind(
  mk(season, 1, "Ashwood", "Eastvale", 10),     mk(season, 2, "Ashwood", "Fairmont", 7),
  mk(season, 3, "Ashwood", "Glenrock", 3),      mk(season, 4, "Harborview", "Ashwood", 6),
  mk(season, 1, "Brightwater", "Glenrock", 14), mk(season, 2, "Brightwater", "Harborview", 21),
  mk(season, 3, "Brightwater", "Ironside", 3),  mk(season, 4, "Dunmore", "Brightwater", 4),
  mk(season, 1, "Cedarcrest", "Ironside", 17),  mk(season, 2, "Cedarcrest", "Dunmore", 9),
  mk(season, 3, "Cedarcrest", "Eastvale", 6),   mk(season, 4, "Fairmont", "Cedarcrest", 2),
  mk(season, 5, "Glenrock", "Cedarcrest", 1),   mk(season, 5, "Dunmore", "Eastvale", 8),
  mk(season, 6, "Fairmont", "Glenrock", 5),     mk(season, 6, "Harborview", "Ironside", 12),
  mk(season, 7, "Eastvale", "Ironside", 4),     mk(season, 7, "Dunmore", "Fairmont", 6)
)

# The conference's published metric. Brightwater rates best; Cedarcrest
# rates above Ashwood.
ssr <- data.frame(
  team = acc_teams$team,
  rating = c(88, 93, 90, 70, 60, 65, 62, 72, 55)
)

rank_season <- function(yr) {
  st <- cfb_standings(acc_games(yr), acc_teams, verbosity = "NONE",
                      tiebreaker_data = list(analytics_ratings = ssr))
  st <- st[order(st$conf_rank), ]
  data.frame(
    Rank = st$conf_rank,
    Team = st$team,
    `Conf W-L` = paste0(st$conf_wins, "-", st$conf_losses),
    `Win pct` = round(st$conf_pct, 3),
    check.names = FALSE
  )[1:5, ]
}
```

```{r past-2025}
show(rank_season(2025), "Ranked under the pre-2026 ACC cascade",
     "The championship game is the top two rows")
```

```{r past-2026}
show(rank_season(2026), "The same games under the 2026 ACC policy",
     "The championship game is the top two rows")
```

**What changed, and why.** Three teams tie at .750 — Ashwood, Brightwater
and Dunmore. Under the pre-2026 cascade the head-to-head chain seats
**Dunmore and Brightwater**. Under the 2026 policy the matchup becomes
**Brightwater and Cedarcrest** — and Cedarcrest is 3-2, a *worse* win
percentage than the three teams above it. It qualifies because it played
five conference games rather than four, so its three wins match the
leaders' three and the candidate-pool rule pulls it into the tie; the SSR
step then rates it above Ashwood. Dunmore, which played for the title under
the old rules, drops out of the top two entirely.

That is not a quirk to route around. It is the policy the ACC adopted, and
the reason it exists is exactly this: teams no longer play the same number
of conference games, so raw win percentage stopped being comparable.

## PRESENT — the 2026 CFP field

The 2026 automatic-qualifier rule is not "the five best champions". It is
the ACC, Big 12, Big Ten and SEC champions **regardless of ranking**, the
**highest-ranked Group-of-6 team whether or not it won its conference**,
and **Notre Dame** when it is ranked inside the field.

```{r present-setup}
board <- data.frame(
  sim = 1L,
  team = c("Ashwood", "Brightwater", "Cedarcrest", "Dunmore", "Eastvale",
           "Fairmont", "Glenrock", "Harborview", "Ironside", "Junction",
           "Kingsport", "Lakeshore", "Marbury", "Northgate", "Oakhurst",
           "Pinehurst", "Quarry Ridge", "Riverton", "Notre Dame", "Summit"),
  conference = c("SEC", "Big Ten", "ACC", "Big 12", "SEC", "Big Ten",
                 "SEC", "Big Ten", "ACC", "Big 12", "Mountain West",
                 "American", "SEC", "Big Ten", "Sun Belt", "MAC",
                 "Big 12", "ACC", "FBS Independents", "Conference USA"),
  conf_champ = c(TRUE, TRUE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE,
                 TRUE, TRUE, TRUE, FALSE, FALSE, FALSE, FALSE, FALSE,
                 FALSE, FALSE, FALSE, FALSE),
  win_pct = seq(1, 0.4, length.out = 20), sov = 0.5, sos = 0.5, pd = 50
)
# The committee's final order. Kingsport - a Group-of-6 CHAMPION - sits at
# 14, one place behind Lakeshore, a Group-of-6 team that is not a champion.
committee <- data.frame(
  team = c("Ashwood", "Brightwater", "Dunmore", "Eastvale", "Fairmont",
           "Glenrock", "Harborview", "Junction", "Marbury", "Northgate",
           "Notre Dame", "Riverton", "Lakeshore", "Kingsport", "Cedarcrest",
           "Ironside", "Oakhurst", "Pinehurst", "Quarry Ridge", "Summit"),
  rank = 1:20
)

field_of <- function(n, policy = "2026") {
  s <- cfb_playoff_seeds(board, rankings = committee, playoff_seeds = n,
                         autobid = policy)
  s <- s[!is.na(s$seed), ]
  s <- s[order(s$seed), ]
  data.frame(
    Seed = s$seed, Team = s$team, Conference = s$conference,
    Champion = ifelse(s$conf_champ, "yes", ""),
    check.names = FALSE
  )
}
```

```{r present-field}
show(field_of(12), "The 12-team field under the 2026 rule",
     "Straight seeding: the field is ordered by committee rank")
```

Two rows in that table are the 2026 rule doing work:

* **Lakeshore (American) is in, and is not a champion.** The Group-of-6
  place goes to the best-ranked G6 *team*, so the American's 13th-ranked
  non-champion takes it.
* **Kingsport, a Group-of-6 champion ranked 14th, misses the field.**
  Outside the Power 4, winning your conference is no longer enough.

Run the same board under the previous era and the difference is explicit:

```{r present-diff}
in26 <- field_of(12)$Team
in25 <- field_of(12, policy = "2025")$Team
data.frame(
  Change = c("In under 2026, out under 2025", "In under 2025, out under 2026"),
  Teams = c(paste(setdiff(in26, in25), collapse = ", "),
            paste(setdiff(in25, in26), collapse = ", "))
) |>
  show("Same board, two automatic-qualifier eras")
```

`autobid = "2025"` reinstates the five-best-champions rule: Kingsport
returns as a champion, Northgate takes the last at-large place, and both
Lakeshore and Notre Dame drop out. **Two of the twelve places change
occupants — four teams affected — on the rule change alone**, without a
single game being played differently. Ten of the twelve are the same either
way, which is the useful part: the eras disagree at the margin, and the
margin is exactly where the arguments happen.

## FUTURE — stress-testing a 16-team field

Nothing in the seeding code is hard-wired to twelve, so raising
`playoff_seeds` answers "who would expansion have let in?" directly:

```{r future-field}
show(field_of(16), "The same board as a 16-team field",
     "The automatic qualifiers are unchanged; the extra places are at-large")
```

```{r future-diff}
added <- setdiff(field_of(16)$Team, field_of(12)$Team)
data.frame(
  `Teams added by expanding 12 to 16` = paste(added, collapse = ", "),
  check.names = FALSE
) |>
  show("What expansion actually buys")
```

The four extra places go to **Cedarcrest, Kingsport, Northgate and
Riverton** — three at-larges plus the Group-of-6 champion who missed the
12-team cut. Note what does *not* change: the automatic qualifiers are the
same teams, because the AQ rule governs who is guaranteed a place, not how
many places exist. Expansion is felt entirely at the bottom of the bracket.

For a simulated 16-team postseason, pass the same argument to
`cfb_simulations()`. The bracket generator handles any field size, giving
byes to the top seeds when the field is not a power of two — 12 seeds
produce the familiar 5v12 / 6v11 / 7v10 / 8v9 first round, 16 produce a
full first round with no byes.

## Feeding the rules real inputs

Several procedures depend on published external numbers that cfbseedR does
not compute. Supply them through `tiebreaker_data`; anything you omit skips
its rung deterministically and records the fact:

```{r inputs, eval = FALSE}
st <- cfb_standings(
  games, teams,
  tiebreaker_data = list(
    analytics_ratings = my_composite,  # SportSource SSR, SP+, SOR, KPI...
    cfp_rankings = latest_cfp,         # the committee's published top 25
    apr = multi_year_apr               # the CUSA policy's late fallback
  )
)
attr(st, "tiebreak_notes")
```

## Simulating, and checking the simulation

Simulations split into `chunks` dispatched with furrr, so a parallel plan
uses every core, and `summary()` renders the result as a gt table:

```{r sim}
games <- cfb_games_example
teams <- cfb_teams_example
games$result[games$week >= 3] <- NA

set.seed(2026)
sim <- cfb_simulations(games, teams, simulations = 1000, playoff_seeds = 4,
                       chunks = 4, verbosity = "NONE")
summary(sim)
```

Three identities make a good smoke test on any simulation output, and this
run satisfies all three exactly:

```{r sim-checks}
data.frame(
  Check = c("playoff probabilities sum to the number of seeds",
            "national-title probabilities sum to one champion",
            "conference titles sum to one per conference"),
  Expected = c(sim$sim_params$playoff_seeds, 1, 2),
  Observed = round(c(sum(sim$overall$playoff),
                     sum(sim$overall$won_natty),
                     sum(sim$overall$conf_champ)), 3)
) |>
  show("Simulation sanity checks", "1000 simulations, seed 2026, 4 chunks")
```

If any of those drifts, something upstream is wrong — a team missing from
`teams`, a conference with no championship game, or a `compute_results`
function that is not filling every scheduled result.

Scaling up is one argument, and progress reporting is one line:

```{r scale, eval = FALSE}
future::plan("multisession")
progressr::handlers(global = TRUE)

sim <- cfb_simulations(games, teams, simulations = 10000, chunks = 8)

future::plan("sequential")
```

A seed reproduces exactly for a fixed `chunks`, and a parallel plan gives
identical results to a sequential one — but the RNG stream depends on the
chunking, so changing `chunks` changes which seasons you draw.

## Pointing this at a real season

Everything above runs on constructed records so the article builds offline.
With cfbfastR installed, the same code takes a real season:

```{r real, eval = FALSE}
sched <- cfbfastR::load_cfb_schedules(2026)
games <- cfb_games_from_schedule(sched)
teams <- cfbfastR::cfbd_team_info(year = 2026) |>
  dplyr::transmute(team = school, conference)

# today's rules on this season
standings <- cfb_standings(games, teams, playoff_seeds = 12)

# the same season re-examined under the previous era's automatic bids
cfb_playoff_seeds(standings, rankings = latest_cfp, autobid = "2025")
```

Because the registry is season-scoped, a `season` column of 2024 or 2025 is
automatically ranked under the rules in force then — so year-over-year
comparisons stay honest with no flags to remember.

## Where the rules come from

Every registered procedure was transcribed from the conference's official
policy (the ACC's July 2026 policy, the SEC/Big Ten/Big 12 2024 policies,
and the published American / CUSA / MAC / Mountain West / Sun Belt
procedures) and the CFP's published formats. The [cheat sheet
(PDF)](https://sportsdataverse.org/cheatsheets/cfbplotR-cfb4th-cfbseedR.pdf)
condenses this package alongside cfbplotR and cfb4th.


# **Our Authors**

-   [Saiem Gilani](https://x.com/saiemgilani)
<a href="https://x.com/saiemgilani" target="blank"><img src="https://img.shields.io/twitter/follow/saiemgilani?color=blue&label=%40saiemgilani&logo=x&style=for-the-badge" alt="@saiemgilani" /></a>
<a href="https://github.com/saiemgilani" target="blank"><img src="https://img.shields.io/github/followers/saiemgilani?color=eee&logo=Github&style=for-the-badge" alt="@saiemgilani" /></a>

## **Adapted from nflseedR (they're awesome)**

cfbseedR is a college-football adaptation of
[nflseedR](https://nflseedr.com) (MIT). The standings-to-seeds architecture,
the week-loop simulator with a pluggable `compute_results`, and the ELO
default generator are all theirs.

-   [Sebastian Carl](https://x.com/mrcaseb)
<a href="https://x.com/mrcaseb" target="blank"><img src="https://img.shields.io/twitter/follow/mrcaseb?color=blue&label=%40mrcaseb&logo=x&style=for-the-badge" alt="@mrcaseb" /></a>
<a href="https://github.com/mrcaseb" target="blank"><img src="https://img.shields.io/github/followers/mrcaseb?color=eee&logo=Github&style=for-the-badge" alt="@mrcaseb" /></a>

-   [Lee Sharpe](https://x.com/LeeSharpeNFL)
<a href="https://x.com/LeeSharpeNFL" target="blank"><img src="https://img.shields.io/twitter/follow/LeeSharpeNFL?color=blue&label=%40LeeSharpeNFL&logo=x&style=for-the-badge" alt="@LeeSharpeNFL" /></a>

## **Citation**

To cite cfbseedR in publications, use:

```
Saiem Gilani (2026). cfbseedR: The SportsDataverse's R Package to Simulate
and Evaluate College Football Seasons.
https://cfbseedR.sportsdataverse.org/
```

```r
citation("cfbseedR")
```

If you use the season-simulation methodology, please also cite
[nflseedR](https://nflseedr.com).

## **Related SportsDataverse packages**

-   [**cfbfastR**](https://cfbfastR.sportsdataverse.org/) — college football
    play-by-play, schedules and rosters; feeds `cfb_games_from_schedule()`
-   [**cfb4th**](https://cfb4th.sportsdataverse.org/) — fourth-down decision
    modeling
-   [**cfbplotR**](https://cfbplotR.sportsdataverse.org/) — team logos and
    plotting helpers for ggplot2
-   [**recruitR**](https://github.com/sportsdataverse/recruitR) — recruiting data
-   [**sportsdataverse-R**](https://r.sportsdataverse.org/) — the meta-package
-   [**sportsdataverse-py**](https://py.sportsdataverse.org/) — the Python
    mirror, whose `cfb_standings` shares this package's tiebreaker rulings

A printable [cheat sheet
(PDF)](https://sportsdataverse.org/cheatsheets/cfbplotR-cfb4th-cfbseedR.pdf)
covers cfbplotR, cfb4th and cfbseedR together.
