---
title: "Simulating seasons"
description: "Custom compute_results functions, the tiebreaker ladder, and CFP-12 seeding."
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article goes deeper than the
[Getting started](https://cfbseedR.sportsdataverse.org/articles/getting-started.html)
walkthrough: writing a custom `compute_results` function, controlling the
tiebreaker cascade with `tiebreaker_depth`, and how the 12-team CFP
straight-seeding rules are applied. Everything runs offline on the bundled
toy season.

```{r setup}
library(cfbseedR)

games <- read.csv(system.file("extdata", "toy_games.csv", package = "cfbseedR"))
teams <- read.csv(system.file("extdata", "toy_teams.csv", package = "cfbseedR"))
games$result[games$week >= 3] <- NA
```

## Custom `compute_results` functions

`cfb_simulations()` calls `compute_results(teams, games, week_num, ...)`
once per remaining week. The contract (identical to nflseedR's):

* return `list(teams = teams, games = games)`,
* fill `result` for exactly the games with
  `week == week_num & is.na(result)`,
* never modify any other result, remove rows/columns from `games`, or
  produce a tie (`result == 0`) outside the regular season.

Here is a deliberately simple generator: every missing result is a home win
by 3 (postseason-safe since 3 is never a tie):

```{r custom-fn}
home_field_rules <- function(teams, games, week_num, ...) {
  fill <- games$week == week_num & is.na(games$result)
  games$result[fill] <- 3
  list(teams = teams, games = games)
}
```

`simulations_verify_fct()` checks a custom function against the full
contract before you burn simulation time on it:

```{r verify}
simulations_verify_fct(home_field_rules)
```

Then plug it in:

```{r custom-sim}
set.seed(1)
sim <- cfb_simulations(
  games, teams,
  compute_results = home_field_rules,
  simulations = 10, playoff_seeds = 4
)
sim$overall[, c("team", "wins", "conf_champ", "playoff")]
```

Anything extra your generator needs (a ratings vector, market lines, an
externally trained model) can be passed through the `...` of
`cfb_simulations()`. The default generator accepts initial ELO ratings that
way:

```{r elo-passthrough}
elo <- setNames(seq(1700, 1300, length.out = nrow(teams)), teams$team)
set.seed(2)
sim_elo <- cfb_simulations(
  games, teams,
  simulations = 10, playoff_seeds = 4,
  elo = elo
)
sim_elo$overall[, c("team", "wins", "playoff")]
```

## The `tiebreaker_depth` ladder

Conference ranks are seeded by conference win percentage; ties are broken
by a cascade whose depth you control:

| `tiebreaker_depth` | Cascade applied before the coin flip |
|---|---|
| `"RANDOM"` | Nothing - coin flip immediately. |
| `"PRE-SOV"` | Head-to-head, then common conference opponents. |
| `"SOS"` (default) | `"PRE-SOV"` + strength of victory, then strength of schedule. |
| `"POINTS"` | `"SOS"` + conference point differential. |

`verbosity = "MAX"` logs every step the cascade takes, which is the easiest
way to understand (and trust) a rank:

```{r tiebreakers}
played <- read.csv(system.file("extdata", "toy_games.csv", package = "cfbseedR"))
standings <- cfb_standings(played, teams,
                           tiebreaker_depth = "POINTS", verbosity = "MAX")
standings[, c("team", "conference", "conf_pct", "sov", "sos", "conf_rank")]
```

### The conference-scoped `sov`/`sos` ruling

All cascade quantities - head-to-head, common opponents, `sov`, `sos`, and
point differential - are computed **over regular-season conference games
only**, so conference ranks depend only on conference play. Concretely:

* `sov` = beaten conference opponents' conference wins divided by their
  conference games (strength of victory over **conference victories**),
* `sos` = all conference opponents' conference wins divided by their
  conference games (strength of schedule over **conference opponents**),
* independents have no conference games, so both are `0.0`.

This is a documented simplification: real CFB tiebreakers are
conference-specific, and a full-schedule SOS would let non-conference
results leak into conference seeding.

### Official per-conference tiebreakers

The generic cascade above is the fallback used for conferences without a
registered procedure. The **SEC, Big Ten, Big 12, ACC, and MAC** instead use
their **official 2024+ tiebreaker procedures** (ported from the same
registry as sdv-py's `cfb_standings.py`, so both engines agree on a shared
cross-language fixture). These add rungs the generic cascade doesn't have -
record vs. common opponents by order of finish, a pooled conference
opponents' win percentage, the SEC's capped relative scoring margin (points
scored capped at 42 / allowed at 48 per game), the Big 12's `total_wins`
with its FCS-or-lower win cap, and an external analytics rating - and they
always run to completion (`tiebreaker_depth` only gates the generic
fallback). Two optional inputs feed these rungs when available:
`home_points`/`away_points` on `games` (the SEC cap) and a `division`
column on `teams` (the Big 12 FCS cap); `tiebreaker_data =
list(analytics_ratings = ratings)` supplies the external rating rung. Any
rung whose input is missing is skipped, and the skip is recorded in
`attr(standings, "tiebreak_notes")`:

```{r registry}
sec_teams <- data.frame(
  team = c("A", "B", "C"), conference = "SEC", division = "FBS"
)
sec_games <- data.frame(
  sim = 2024, week = 1:3, game_type = "REG",
  home_team = c("A", "B", "C"), away_team = c("B", "C", "A"),
  result = c(50, 3, 3), neutral = 0,
  home_points = c(50, 20, 20), away_points = c(0, 17, 17)
)
sec_standings <- cfb_standings(sec_games, sec_teams,
                               tiebreaker_depth = "POINTS", verbosity = "NONE")
sec_standings[, c("team", "conf_rank")]
attr(sec_standings, "tiebreak_notes")
```

## CFP-12 straight seeding

`cfb_playoff_seeds()` implements the 2025 **straight seeding** rule:

1. Order all teams by committee rank (`rankings`); unranked teams fall back
   behind ranked ones (and to win pct, SOV, SOS, and point differential
   when `rankings = NULL`).
2. The 5 highest-ranked **conference champions** are guaranteed a spot
   (fewer if fewer champions exist).
3. Fill the remaining spots with the best-ranked at-large teams.
4. Seed the field strictly in ranking order - champions are **not** bumped
   up to the top-4 seeds (that was the pre-2025 rule).

A synthetic 16-team example with 6 conference champions shows the
guarantee and the straight ordering; champion `F1` (ranked 14th) displaces
the lowest at-large team but keeps its rank-order seed:

```{r cfp12}
st <- data.frame(
  sim = 1,
  team = sprintf("%s1", LETTERS[1:16]),
  conference = c(LETTERS[1:6], LETTERS[1:10]),
  conf_champ = c(rep(TRUE, 6), rep(FALSE, 10)),
  win_pct = seq(1, 0.4, length.out = 16),
  sov = 0.5, sos = 0.5, pd = 100
)
rankings <- data.frame(
  team = st$team,
  rank = c(1, 2, 3, 5, 8, 14, 4, 6, 7, 9, 10, 11, 12, 13, 15, 16)
)
seeded <- cfb_playoff_seeds(st, rankings = rankings, playoff_seeds = 12)
seeded[!is.na(seeded$seed), c("team", "conf_champ", "seed")] |>
  (\(x) x[order(x$seed), ])()
```

In `cfb_simulations()`, the bracket generated from these seeds is the fixed
CFP-12 bracket (first round hosted by the higher seed, later rounds
neutral-site, no reseeding), and committee `rankings` are held static
across simulations.

## Real schedules

With cfbfastR installed, a real season plugs straight in:

```{r real, eval = FALSE}
# Not run: requires network access via cfbfastR
sched <- cfbfastR::load_cfb_schedules(2024)
games_2024 <- cfb_games_from_schedule(sched)
games_2024 <- games_2024[games_2024$game_type != "POST", ]
teams_2024 <- unique(data.frame(
  team = c(sched$home_team, sched$away_team),
  conference = c(sched$home_conference, sched$away_conference)
))
sim <- cfb_simulations(games_2024, teams_2024, simulations = 1000)
```
