---
title: "Cookbook: tiebreakers & the 2026 CFP rules"
description: >
  Recipes for the official per-conference tiebreakers, division formats,
  external metric inputs, and the 2026 CFP automatic-qualifier policy.
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

cfbseedR implements each FBS conference's **official championship-game
tiebreaker procedure**, registered as season-scoped epochs, and the CFP's
**season-keyed automatic-qualifier policies**. This cookbook shows how the
pieces fit together, one recipe at a time. Everything below runs offline.

```{r setup}
library(cfbseedR)
```

## The rules, at a glance (2026 season)

| Conference | Format | Procedure (after win pct) |
|---|---|---|
| SEC | single table | h2h → common opponents → standings descent → opponents' win pct → capped scoring margin (42/48) → draw |
| Big Ten | single table | h2h → common opponents → descent → opponents' win pct → analytics rating → draw |
| Big 12 | single table | h2h → common opponents → descent → opponents' win pct → capped total wins → analytics → draw |
| ACC (2026+) | single table | candidate pool by wins-or-losses → h2h → SportSource SSR → draw |
| American | single table | h2h → CFP-ranked final-week clause → metric composite → common opponents → draw |
| Conference USA | single table | h2h → CFP-ranked clause → metric composite → APR → draw |
| MAC | single table | h2h → common opponents → SportSource rating → draw |
| Mountain West | single table | metric composite → h2h → draw |
| Sun Belt | **two divisions** | h2h → divisional win pct → descent → common opponents → CFP-ranked clause → composite → draw |
| Pac-12 (re-formed) | single table | no published procedure — generic fallback |

The CFP field (12 teams, straight seeding) admits the ACC / Big 12 /
Big Ten / SEC champions regardless of ranking, the highest-ranked
Group-of-6 team (champion or not), and Notre Dame when ranked inside the
field (`autobid = "2026"`, the default). `autobid = "2025"` restores the
five-best-champions rule used in 2024-2025.

## Recipe 1: season-scoped epochs

The same schedule ranked as a 2025 season and as a 2026 season can crown
different ACC championship-game participants, because the ACC replaced its
entire procedure in 2026. Supply a `season` id column and the registry
resolves the right epoch; a plain `sim` id always means "current rules".

```{r epochs}
teams <- data.frame(team = c("A", "B", "C", "D", "E"), conference = "ACC")
games <- data.frame(
  season = 2026, week = c(1, 2, 1, 2, 3, 4, 3), game_type = "REG",
  home_team = c("A", "A", "B", "B", "C", "C", "D"),
  away_team = c("C", "D", "C", "E", "D", "E", "E"),
  result = 7
)
ratings <- data.frame(team = c("A", "B", "C"), rating = c(1, 2, 0.5))

st <- cfb_standings(games, teams, verbosity = "NONE",
  tiebreaker_data = list(analytics_ratings = ratings)
)
st[, c("team", "conf_rank", "conf_pct")]
```

Change `games$season` to `2025` and the pre-2026 cascade (common
opponents, standings descent, opponents' win pct) applies instead.

## Recipe 2: external metric inputs

The American, CUSA, Mountain West, and Sun Belt procedures lean on
published metric composites (Connelly SP+, ESPN SOR, KPI, SportSource) and
the CFP committee rankings. cfbseedR computes none of these — you supply
them through `tiebreaker_data`, and any missing input skips its rung
deterministically (recorded in `attr(result, "tiebreak_notes")`):

```{r inputs, eval = FALSE}
st <- cfb_standings(
  games, teams,
  tiebreaker_data = list(
    # your composite of the conference's published metrics
    analytics_ratings = data.frame(team = teams$team, rating = my_composite),
    # the latest CFP committee rankings
    cfp_rankings = data.frame(team = c("Tulane", "Navy"), rank = c(24, 25)),
    # multi-year APR (the CUSA policy's late fallback)
    apr = data.frame(team = teams$team, apr = my_apr)
  )
)
attr(st, "tiebreak_notes")
```

The `cfp_ranked_final_week` clause is faithful to the published text: the
best-ranked tied team advances only if it **won** its final conference
game; a ranked team that lost falls through to the composite.

## Recipe 3: division formats (Sun Belt)

Add a `conf_division` column to `teams` and that conference is ranked
division-first: each division runs the official procedure internally and
the division champions take `conf_rank` 1-2 (the championship-game
matchup). This is the 2026 Sun Belt — and equally any historical
divisional season you want to re-rank.

```{r divisions}
sb_teams <- data.frame(
  team = c("E1", "E2", "W1", "W2"),
  conference = "Sun Belt",
  conf_division = c("East", "East", "West", "West")
)
sb_games <- data.frame(
  season = 2026, week = c(1, 1, 2, 2), game_type = "REG",
  home_team = c("E1", "W1", "E1", "E2"),
  away_team = c("E2", "W2", "W1", "W2"),
  result = 7
)
sb <- cfb_standings(sb_games, sb_teams, verbosity = "NONE")
sb[, c("team", "conference", "conf_rank")]
```

`W1` (1-1) outranks `E2` (1-1) because divisions, not the overall table,
decide who reaches the championship game.

## Recipe 4: the 2026 CFP field

```{r cfp}
st <- data.frame(
  sim = 1L,
  team = c("BigTenChamp", "AccChamp", "B12Champ", "SecChamp",
           "G6Team", "Notre Dame", paste0("AtLarge", 1:6)),
  conference = c("Big Ten", "ACC", "Big 12", "SEC",
                 "Mountain West", "FBS Independents", rep("Big Ten", 6)),
  conf_champ = c(TRUE, TRUE, TRUE, TRUE, FALSE, FALSE, rep(FALSE, 6)),
  win_pct = seq(1, 0.45, length.out = 12), sov = 0, sos = 0, pd = 0
)
rankings <- data.frame(team = st$team, rank = seq_len(12))

seeded <- cfb_playoff_seeds(st, rankings = rankings, playoff_seeds = 12)
seeded[order(seeded$seed), c("team", "conference", "seed")]
```

Note the Mountain West team enters as the Group-of-6 automatic qualifier
without winning its conference, and an unranked Power-4 champion would
still make the field. Straight seeding means nobody is bumped up: seeds
follow the committee order exactly, with the top four seeds earning
first-round byes in `cfb_simulations()`.

## Recipe 5: scaling up

Ten thousand seasons is a `chunks` argument away. Simulations are split into
contiguous blocks and dispatched with furrr, so a parallel plan uses every
core; `summary()` renders the result as a gt table grouped by conference.

```{r scale, eval = FALSE}
future::plan("multisession")
progressr::handlers(global = TRUE)

sim <- cfb_simulations(
  games, teams,
  simulations = 10000,
  chunks = 8,
  autobid = "2026",
  tiebreaker_data = list(analytics_ratings = my_composite)
)
summary(sim)

future::plan("sequential")
```

A seed reproduces exactly for a fixed `chunks`, and parallel matches
sequential -- but the stream depends on the chunking, so changing `chunks`
changes which seasons you get.

## Where the rules come from

Every registered procedure was transcribed from the conference's official
policy (ACC July 2026 policy, SEC/Big Ten/Big 12 2024 policies, the
published American/CUSA/MAC/Mountain West/Sun Belt procedures) and the
CFP's published formats. The [cheat sheet
(PDF)](https://sportsdataverse.org/cheatsheets/cfbplotR-cfb4th-cfbseedR.pdf)
condenses this package alongside cfbplotR and cfb4th.
